### 2014-10-22 ApTranscriptome_TR.Rmd
updated genetic assimilation analysis and figure

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice Bacon DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*. For historical reasons, I refer to these species as colonies at times throughout this technical report.

@@

@@

chi1 <- chisq.test(resp.table)

# test if the number of transcripts in each category differs from null expectations using Pearson Chi-square test
chi2 <- chisq.test(Ap.type.table)
chi2
```

@@

# A. carolinensis **Low** and **Bimodal**
md_sub1 <- subset(md, Acar %in% c("Low", "Bimodal"), )
md_sub1 <- droplevels(md_sub1)

mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") +
  theme(legend.position = "none")
mh_plot_sub1

md_sub2 <- subset(md, Apic %in% c("High", "Bimodal"), )
md_sub2 <- droplevels(md_sub2)

mh_plot_sub2 <- qplot(x=Apic, y=Acar, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") + 
  theme(legend.position = "none")
mh_plot_sub2

# get legend from http://stackoverflow.com/questions/11883844/inserting-a-table-under-the-legend-in-a-ggplot2-histogram
#Extract Legend
g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

legend <- g_legend(mh_plot)

# remove legend from mh_plot
mh_plot <- mh_plot + theme(legend.position = "none")

# add bars
geom_rect(data = dark.subset, aes(xmin = phase_start, xmax = phase_end), 
              ymin = -Inf,
              xmax = Inf,
              fill = alpha("#000000", 0.2))

# arrange plots

grid.arrange(mh_plot, mh_plot_sub2, mh_plot_sub1, legend,
             ncol=2, widths=c(1,0.5), heights=c(1, 0.5))

arrangeGrob(mh_plot, mh_plot_sub1, mh_plot_sub2, ncol=1), 
                    ncol=2, widths=c(1,1.2))

tiff("results/matched_observations_Fig1.tiff", compression = "lzw")
mh_plot
dev.off()
```

@@

An interesting observation from the matched observations plot (Fig. ...) is that for **NotResp** genes in *A. carolinensis* the **High** and **Bimodal** categories in *A. picea* are over-represented. Finding that the mean expression level of these genes in *A. carolinensis* is greater than the expression level at the optimum temperature (19.5°C) in *A. picea* would be consistent with genetic assimilation.

@@

To visualize this data, I plot the log ratio of inducibility between species against the log ratio of baseline expression between species. I do this separately for **High** transcripts in *A. picea* and then **Low** transcripts in *A. carolinensis*.

@@

I extract the **High** transcripts for *A. picea* and plot the log ratio of inducibility in *A. picea* over inducibility in *A. carolinensis*, against the log ratio of baseline expression in *A. carolinensis* against expression in *A. picea*. According to the *genetic assimilation hypothesis* there should be a positive relationship between these ratios. That is, greater baseline expression in *A. carolinensis* should have greater inducibility towards **High** temperatures in *A. picea*.

@@

# ratio of baseline expression
Ap.response.type2.A22.high$Ar.A22.baseline <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt

# plot!
gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg+0.01), x = log(Ar.A22.baseline))) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(y = expression(paste("log(inducibility ", italic("A. picea / A. carolinensis"), ")")),
       x = expression(paste("log(baseline ", italic("A. carolinensis / A. picea"), ")")))    
gg11

A22.high.lm <- lm(log(A22.Ar.upreg) ~ log(Ar.A22.baseline), data = Ap.response.type2.A22.high)

@@

In contrast to the prediction there is a negative relationship between these ratios. That is, genes with greater upregulation of expression also have greater baseline expression. This is also consistent with the results simply comparing baseline expression between species for the *A. picea* **High** transcripts.

@@

```{r assimilation_Ar_low}
### Repeat for A. carolinensis *Low* transcripts

Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
str(Ap.response.type2.Ar.low)

@@

# ratio of baseline expression
Ap.response.type2.Ar.low$A22.Ar.baseline <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt

gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.baseline))) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(y = expression(paste("log(inducibility ", italic("A. carolinensis / A. picea"), ")")),
       x = expression(paste("log(baseline ", italic("A. picea / A. carolinensis"), ")")))

@@

Ar.low.lm <- lm(log(Ar.A22.upreg) ~ log(A22.Ar.baseline), data = Ap.response.type2.Ar.low)

@@

```{r fig3}
# grid.arrange
png(file = "results/genetic_assimilation.png")
grid.arrange(gg11, gg12, nrow = 1)
dev.off()
```

@@

As with genes turned on at high temperatures, *A. carolinensis* **Intermediate** genes are down-regulated on average at 2.7°C higher temperatures than *A. picea* **Intermediate** genes. In contrast, towards low temperatures, the temperature of mean temperature of downregulation was higher for *A. picea* (10.6°) than *A. carolinensis* (9.75°).

@@

# grid.arrange
#tiff("results/crit_temp_regulation_Fig2.tiff", compression = "lzw", bg = "white")
grid.arrange(gg2, gg3, nrow = 2)
#dev.off()
```

>>>>

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*. For historical reasons, I sometimes refer to these species as colonies at times throughout this technical report.

@@

# test if the number of transcripts in each category differs from null expectations using Pearson Chi-square test
(chi1 <- chisq.test(Ap.type.table))

@@

(chi2 <- chisq.test(resp.table))

# test if the number of Low responsive transcripts differs between species 
low.table <- matrix(nrow=2, ncol=2, dimnames = list(c("Low", "NotLow"), c("Ap", "Ac")))
low.table[1,1] <- A22.type.table[which(names(A22.type.table) == "Low")]
low.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "Low")])
low.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "Low")]
low.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "Low")])
low.table

(chi3 <- chisq.test(low.table))

# test if the number of Intermediate responsive transcripts differs between species 
intermediate.table <- matrix(nrow=2, ncol=2, dimnames = list(c("Intermediate", "Others"), c("Ap", "Ac")))
intermediate.table[1,1] <- A22.type.table[which(names(A22.type.table) == "Intermediate")]
intermediate.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "Intermediate")])
intermediate.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "Intermediate")]
intermediate.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "Intermediate")])
intermediate.table

(chi4 <- chisq.test(intermediate.table))

# test if the number of High responsive transcripts differs between species 
high.table <- matrix(nrow=2, ncol=2, dimnames = list(c("High", "Others"), c("Ap", "Ac")))
high.table[1,1] <- A22.type.table[which(names(A22.type.table) == "High")]
high.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "High")])
high.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "High")]
high.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "High")])
high.table

(chi5 <- chisq.test(high.table))

# test if the number of Bimodal responsive transcripts differs between species 
bimodal.table <- matrix(nrow=2, ncol=2, dimnames = list(c("Bimodal", "Others"), c("Ap", "Ac")))
bimodal.table[1,1] <- A22.type.table[which(names(A22.type.table) == "Bimodal")]
bimodal.table[2,1] <- sum(A22.type.table[which(names(A22.type.table) != "Bimodal")])
bimodal.table[1,2] <- Ar.type.table[which(names(Ar.type.table) == "Bimodal")]
bimodal.table[2,2] <- sum(Ar.type.table[which(names(Ar.type.table) != "Bimodal")])
bimodal.table

(chi6 <- chisq.test(bimodal.table))
```

@@

# A. carolinensis **Low** and **Bimodal**
md_sub1 <- subset(md, Acar %in% c("Low", "Bimodal", "High"), )
md_sub1 <- droplevels(md_sub1)

mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") +
  theme(legend.position = "bottom") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = .5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

mh_plot_sub1

md_sub2 <- subset(md, Apic %in% c("Low", "High", "Bimodal"), )
md_sub2 <- droplevels(md_sub2)

mh_plot_sub2 <- qplot(x=Acar, y=Apic, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Apic))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. carolinensis", y = "A. picea") + 
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

mh_plot_sub2

# arrange plots to make figure for manuscript
mh_plot_sub2 <- mh_plot_sub2 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
mh_plot_sub1 <- mh_plot_sub1 + ggtitle('B') + theme(plot.title = element_text(hjust=0))

png("results/matched_observations_Fig1.png")
grid.arrange(mh_plot_sub2, mh_plot_sub1, ncol=1, heights = c(0.45, 0.6))
dev.off()

# full figure for supplemental
png("results/matched_observations_FigS1.png")
mh_plot
dev.off()
```

@@

An interesting observation from the matched observations plot (Fig. 1) is that for **NotResp** genes in *A. carolinensis* the **High** and **Bimodal** categories in *A. picea* are over-represented. Finding that the mean expression level of these genes in *A. carolinensis* is greater than the expression level at the optimum temperature (19.5°C) in *A. picea* would be consistent with genetic assimilation.

@@

To visualize this data, I plot the log ratio of inducibility between species against the log ratio of constitutive expression between species. I do this separately for **High** transcripts in *A. picea* and then **Low** transcripts in *A. carolinensis*.

@@

I extract the **High** transcripts for *A. picea* and plot the log ratio of inducibility in *A. picea* over inducibility in *A. carolinensis*, against the log ratio of constitutive expression in *A. carolinensis* against expression in *A. picea*. According to the *genetic assimilation hypothesis* there should be a positive relationship between these ratios. That is, greater constitutive expression in *A. carolinensis* should have greater inducibility towards **High** temperatures in *A. picea*.

@@

# ratio of constitutive expression
Ap.response.type2.A22.high$Ar.A22.constitutive <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt

# plot!
gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg), x = log(Ar.A22.constitutive))) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(y = expression(paste("log(inducibility ", italic("A. picea / A. carolinensis"), ")")),
       x = expression(paste("log(constitutive ", italic("A. carolinensis / A. picea"), ")")))

gg11

A22.high.lm <- lm(log(A22.Ar.upreg) ~ log(Ar.A22.constitutive), data = Ap.response.type2.A22.high)

@@

In contrast to the prediction there is a negative relationship between these ratios. That is, genes with greater upregulation of expression also have greater constitutive expression. This is also consistent with the results simply comparing constitutive expression between species for the *A. picea* **High** transcripts.

@@

```{r assimilation_Ar_low}
### Repeat for A. carolinensis *Low* transcripts
Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
str(Ap.response.type2.Ar.low)

@@

# ratio of constitutive expression
Ap.response.type2.Ar.low$A22.Ar.constitutive <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt

gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.constitutive))) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(y = expression(paste("log(inducibility ", italic("A. carolinensis / A. picea"), ")")),
       x = expression(paste("log(constitutive ", italic("A. picea / A. carolinensis"), ")")))

@@

Ar.low.lm <- lm(log(Ar.A22.upreg) ~ log(A22.Ar.constitutive), data = Ap.response.type2.Ar.low)

@@

```{r fig2}
# add labels
gg11 <- gg11 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
gg12 <- gg12 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
# grid.arrange
png(file = "results/genetic_assimilation_Fig2.png")
grid.arrange(gg11, gg12, nrow = 1)
dev.off()
```

@@

As with genes turned on at high temperatures, *A. carolinensis* **Intermediate** genes are down-regulated on at higher temperatures than *A. picea* **Intermediate** genes. In contrast, towards low temperatures, the temperature of mean temperature of downregulation was higher for *A. picea* than *A. carolinensis*.

@@

# grid.arrange
gg2 <- gg2 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
gg3 <- gg3 + ggtitle('B') + theme(plot.title = element_text(hjust=0))

png("results/crit_temp_regulation_Fig3.png")
grid.arrange(gg2, gg3, nrow = 2, heights = c(.5,.6))
dev.off()
```
